---
title: "random-coefficients-model"
author: "Katy"
date: "2/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # data manipulation and graphics
library(lubridate) # manipulation of date data
library(lme4) # fitting general(ized) linear mixed models
library(car) # Anova function for Type III tests  !!!NB!!! car masks dplyr::recode
library(emmeans) # estimate means and slopes, rough interaction plots
library(DHARMa) # checking assumptions of mixed model
library(cAIC4) # conditional AIC for model comparison
```

Hi Susan -
I've taken the code you provided for me and used it below, and will annotate where I have issues or questions.
```{r}

# import data----
head.wide <- read.csv("../processed_data/head_growth_wide.csv")
# note the flat and parsimonious structure of this file

# restructure from wide to long----
head.long <- gather(head.wide, sdate, head_width, na.rm = F, X28.Jul:X28.Dec)

# recode sdate to sdate2 as character class----
head.long <- head.long %>%
  mutate(sdate2 = dplyr::recode(sdate,                     # note use of dplyr::
                                X28.Jul  = "2020-07-28",
                                X10.Aug = "2020-08-10",
                                X24.Aug = "2020-08-24",
                                X8.Sep = "2020-09-08",
                                X21.Sep = "2020-09-21",
                                X5.Oct = "2020-10-05",
                                X19.Oct = "2020-10-19",
                                X2.Nov = "2020-11-02",
                                X16.Nov = "2020-11-16",
                                X30.Nov = "2020-11-30",
                                X14.Dec = "2020-12-14",
                                X28.Dec = "2020-12-28"))

# check recoding----
with(head.long, table(sdate2, sdate))

# create sampling_date (date class) from sdate2 (character class)----
head.long <- head.long %>%
  mutate(sampling_date = lubridate::ymd(sdate2))

# date data are stored as integers: the number of days since 1960-01-01 minus 1
# see https://stats.idre.ucla.edu/r/faq/how-does-r-handle-date-values/ 

# create rescaled date, such that 2019-07-01 is equivalent to zero, to use in regression----
# for a nice discussion of this use of rescaling, see 
#   https://stats.stackexchange.com/questions/65900/does-it-make-sense-to-use-a-date-variable-in-a-regression

# sampling_date_scaled is number of days since 2020-07-28
head.long<- head.long %>%
  mutate(sampling_date_scaled = sampling_date - as.Date("2020-07-28"))

# plot profile for each individual in each treatment----
ggplot(head.long, aes(x = sampling_date, y = head_width, color=ID.)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ TANK) +
  theme(legend.position = "none")

# use smoother to get a sense of the shape of the relationship----
ggplot(head.long, aes(x = sampling_date, y = head_width, color=ID.)) +
  geom_smooth(se = F, span=2) +
  geom_point() +
  facet_wrap(~ TANK) +
  theme(legend.position = "none") # turns legend off

```

##ASK SUSAN: What is the purpose of the line of code below? 

```{r}
## random coefficients model
# make a copy of the data frame used in the model, to make life easier
df <- filter(head.long, sampling_date <= as.Date("2020-12-28") & !is.na(head_width))

```

```{r}
# convert TANK to factor to work in model below

df['TANK'] <- lapply(df['TANK'], factor)

# fit random intercepts only model----
# slopes for all ID. within a TANK will be equal
m1 <- lmer(head_width ~ TANK * sampling_date_scaled + (1 | ID.),
           data = df,
           contrasts = list(TANK = contr.sum)) #need to convert TANK to factor for this code to work
summary(m1)
car::Anova(m1)

simulationOutput <- simulateResiduals(fittedModel = m1)

simulationOutput
testUniformity(simulationOutput)

```

## ASK SUSAN: deviation significant here, what does that mean?

```{r}
# plot scaled residuals versus fitted to check homogeneity of variance----
plotResiduals(simulationOutput$fittedPredictedResponse, simulationOutput$scaledResiduals)
```

##ASK SUSAN: why does my plot look so different than hers? What does it mean?

```{r}
## plot scaled residuals versus Trt to check homogeneity of variance----
plotResiduals(df$TRTMENT, simulationOutput$scaledResiduals)
## above plot looks wrong

# plot scaled residuals versus sampling_date_scaled to check homogeneity of variance----
# plotResiduals(df$sampling_date_scaled, simulationOutput$scaledResiduals)
##why won't this line of code work (above)

# DHARMa also has a testTemporalAutocorrelation function that could be useful
# but the code is complicated when there is more than one observation at each time
# I haven't done it, but you can :)


# capture population-averaged regressions----
df$yhat_pa <- predict(m1, re.form = NA)
#### ideally add ci
# see https://datascienceplus.com/introduction-to-bootstrap-with-applications-to-mixed-effect-models/

# capture regressions for each Cup----
df$yhat <- predict(m1, re.form = NULL)

# plot observations, population-averaged regressions and cup-level regressions----
ggplot(data=df, ) +
  geom_point(aes(x = sampling_date, y = head_width, color = ID.)) +
  geom_line(aes(x = sampling_date, y = yhat_pa)) +
  geom_line(aes(x = sampling_date, y = yhat, color = ID.), linetype = "dashed") +
  facet_wrap(~ TANK) +
  theme(legend.position = "none")

# fit model with random intercepts and random slopes, uncorrelated----
# slopes for ID. within a Trt will vary
m2 <- lmer(head_width ~ TANK * sampling_date_scaled + (sampling_date_scaled || ID.),
           
           data = df,
           contrasts = list(TANK = contr.sum),
           control = lmerControl(optimizer ="bobyqa"))

summary(m2)
```

#ASK SUSAN: error regarding very large eigenvalue ?

```{r}
# check assumptions
## NEED TO ADD CODE CHECKING ASSUMPTIONS


# compute population-averaged regressions----
df2 <- df
df2$yhat_pa <- predict(m2, re.form = NA)

# compute regressions for each Cup----
df2$yhat <- predict(m2, re.form = NULL)

# plot observations, population-averaged regressions and cup-level regressions----
# it is OK to model using sampling_date_scaled and to plot using sampling_date
ggplot(data=df2) +
  geom_point(aes(x = sampling_date, y = head_width, color = ID.)) +
  geom_line(aes(x = sampling_date, y = yhat_pa)) +
  geom_line(aes(x = sampling_date, y = yhat, color = ID.), linetype = "dashed") +
  facet_wrap(~ TANK) +
  theme(legend.position = "none")

# plot observations, population-averaged regressions and cup-level regressions----
# with bootstrapped 95% CI on regression
# CIs in mixed models in R are not easy to obtain; I cannot vouch for these but they look OK
source("bootpredictlme4.R", print = T)
## Registered S3 method overwritten by 'bootpredictlme4':
## method from
## predict.merMod lme4

```

##ASK SUSAN: what is the above line of code supposed to do? doesn't print anything, takes a while to run.

```{r}
# extract regression parameter estimates and SEs for each Trt regression----
# this is the same model as m2, just reparameterized
#m2b <- lmer(length ~ 0 + TANK + TANK : sampling_date_scaled + (sampling_date_scaled || ID.),
            
            data = df2,
            contrasts = list(TANK = contr.sum),
            control = lmerControl(optimizer ="bobyqa"))

#summary(m2b)
```

#do not have above code working, "error in model.frame.default(data = df2, drop.unused.levels = TRUE, formula = length ~  : 
#object is not a matrix"


```{r}

# pairwise comparisons of slopes----
#(lst <- emtrends(m2, "TANK", var = "as.numeric(sampling_date_scaled)"))

# produces the same estimates for slopes ("trend") as model m2b
#(co <- contrast(lst, method = "pairwise"))

# compare model without random slopes to model with random slopes----
# using a likelihood ratio test (or AIC)
#anova(m1, m2)

# using Safken et al
# Conditional Model Selection in Mixed-Effects Models with cAIC4
# arXiv:1803.05664v2 [stat.CO] 17 Mar 2018
#cAIC(m1)
#cAIC(m2)

```



